Optimizing the power of adaptive seamless designs using Bayesian Optimization

### 2020-09-17 main.tex
joerg changes

The strategies regarding treatment selection for the second stage in the two-phase trial design, following the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}, and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of the selection strategies that determine which trial arms are kept for the second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation & Procedure of selection strategy  \\
  \hline
  \emph{all}     & Select all arms  \\
  $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than \emph{eps} (epsilon) below maximal test statistic \\
  \emph{thresh}     & Select all arms with test statistic larger than the value \emph{thresh} (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter         & Range \\
  \hline
  Selection strategy & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thr} \}$ \\
  $r$ (ratio)        & $(0,1)$ \\
  eps (epsilon)            & $[0,4]$ \\
  thresh (threshold)          & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

@@

The total number of treatments $n_{\text{treat}}$ should be constant in the simulation study, when comparing the efficiency of different designs.
We denote the proportion of patients per treatment in stage 1, compared to stage 2, with 
$$ r=\frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}$$
To keep the number of total treatments constant we optimize the ratio ($r \in (0,1)$) of $n_{\text{stage1}}$ and $n_{\text{stage2}}$ instead of optimizing these values directly.
Therefore we define
\begin{align}
  \label{eq:stagec}
  %\begin{split}
  n_{\text{stage1}} &= r \cdot c, \qquad
  n_{\text{stage2}} = (1-r) \cdot c,
  %\end{split}
\end{align}
whereas $c$ is a constant that depends on $n_{\text{treat}}$, $k_1$ and $k_2$.
For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
For a general solution, we can insert the variables from (\ref{eq:stagec}) into (\ref{eq:ntreat}) and obtain
\begin{align}
 % \label{eq:ntreatc}
  n_{\text{treat}} & = k_1 \cdot r \cdot c + k_2 \cdot (1-r) \cdot c \;\; \Leftrightarrow \;\; %\\
  \label{eq:ntreatcc}
  c = \frac{n_{\text{treat}}}{k_1 \cdot r + k_2 \cdot (1-r)}.
\end{align}
Placing $c$ into (\ref{eq:stagec}) lets us calculate $n_{\text{stage1}}$ and $n_{\text{stage2}}$ for the given constraints and a given ratio $r$.
This formula gives us the treatments per arm and per stage if $k_2$ is known. 
Unfortunately, this is only the case for \emph{selection strategy} $\in \{\text{all},1,2,3\}$.

For the selection strategies \emph{thr} and \emph{eps}, the number of arms in the second stage ($k_2$) depends on the choice of the \emph{Threshold} or \emph{Epsilon} parameter.
However, for given values of $n_{\text{stage1}}$ and \emph{Threshold} or \emph{Epsilon} respectively we can estimate the expected number of arms at stage 2 ($\hat{k}_2$), by running a simulation of the first stage.
This simulation is run multiple times for different values of $n_{\text{stage1}}$ in a calibration step until the desired number of $n_{\text{treat}}$ is reached approximately.
In fact, this calibration step formulates an independent optimization problem:
We minimize 
\begin{align}
  \label{eq:targettreat}
  \begin{split}
  & \operatorname{min}_{n_{\text{stage1}} \in [l,u]} h(n_{\text{stage1}}) \,  \text{, with} \\
  h(n_{\text{stage1}}) &:= \biggl( (k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot \underbrace{\frac{1-r}{r} \cdot n_{\text{stage1}}}_{=n_{\text{stage2}}}) - n_{\text{treat}} \biggr)^2 \, \text{and} \\
  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
  \end{split}
\end{align}
and $l = \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil$ and $u = \lceil \frac{n_{\text{treat}}}{k_1} \rceil$. 
$f$ is the simulator black-box from Equation~\ref{eq:bbox}, that gives us the estimated number of $\hat{k}_2$.
This number is independent of the simulation of the second stage, so we can set $n_{\text{stage2}}$ to 1.
The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.

>>>>

The strategies regarding treatment selection for the second stage in the two-phase trial design, following the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}, and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation & Procedure of selection strategy  \\
  \hline
  \emph{all}     & Select all arms  \\
  $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than \emph{eps} (epsilon) below maximal test statistic \\
  \emph{thresh}     & Select all arms with test statistic larger than the value \emph{thresh} (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter         & Range \\
  \hline
  Selection strategy & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thr} \}$ \\
  $r$ (ratio)        & $(0,1)$ \\
  eps (epsilon)            & $[0,4]$ \\
  thresh (threshold)          & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

@@

The total number of treatments $n_{\text{treat}}$ should be constant in the simulation study, when comparing the efficiency of different designs.
We denote the proportion of patients per treatment in stage 1, compared to patients per treatment in both stages, with 
$$ r=\frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}.$$
Given the number of treatments $k_1$ and $k_2$ and the total number of patients $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
\begin{align}
  \label{eq:stagec}
  %\begin{split}
  n_{\text{stage1}} & = 
  n_{\text{stage1}}\cdot \frac{n_{\text{treat}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} = 
  n_{\text{treat}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \\
  & = n_{\text{treat}}\cdot
  \frac{r}{k_1\cdot
  r+k_2\cdot (1-r)}, \\
  n_{\text{stage2}} & =
  n_{\text{treat}}\cdot
  \frac{1-r}{k_1\cdot
  r+k_2\cdot (1-r)}.
  %\end{split}
\end{align}
%For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
These formulas give the treatments per arm and per stage if $k_2$ is known. 
However, this is only the case for selection strategies 1, 2, and 3.
For the selection strategies \emph{threshold} and \emph{epsilon}, the number of arms in the second stage ($k_2$) depends on the choice of the \emph{thresh} or \emph{eps} parameter.
However, for given values of $n_{\text{stage1}}$ and \emph{Threshold} or \emph{Epsilon} respectively we can estimate the expected number of arms at stage 2 ($\hat{k}_2$), by running a simulation of the first stage.
This simulation is run multiple times for different values of $n_{\text{stage1}}$ in a calibration step until the desired number of $n_{\text{treat}}$ is reached approximately.
In fact, this calibration step formulates an independent optimization problem:
We minimize 
\begin{align}
  \label{eq:targettreat}
  \begin{split}
  & \operatorname{min}_{n_{\text{stage1}} \in [l,u]} h(n_{\text{stage1}}) \,  \text{, with} \\
  h(n_{\text{stage1}}) &:= \biggl( (k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot \underbrace{\frac{1-r}{r} \cdot n_{\text{stage1}}}_{=n_{\text{stage2}}}) - n_{\text{treat}} \biggr)^2 \, \text{and} \\
  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
  \end{split}
\end{align}
and $l = \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil$ and $u = \lceil \frac{n_{\text{treat}}}{k_1} \rceil$. 
$f$ is the simulator black-box from Equation~\ref{eq:bbox}, that gives us the estimated number of $\hat{k}_2$.
This number is independent of the simulation of the second stage, so we can set $n_{\text{stage2}}$ to 1.
The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.
